---
title: "Colour Game"
format: html
---
::: tag-list
[Permutation Test]{.tag} [Bootstrap]{.tag} [Between-Subjects Design]{.tag} [Resampling]{.tag}
:::

## About

This app is a fast-paced colour-matching game, structurally similar to the Stack Game but with a different task and a different outcome measure. As before, are assigned to one of two conditions before playing: **Control**, the standard game with no interruptions, or **Distraction**, where notification-style pop-ups appear periodically during play. Note: if following on from the Stack game, you could switch groups.

Everyone plays the same underlying game, only the presence of distractions differs between groups, making this another clean between-subjects comparison - this time used to introduce **resampling-based** hypothesis testing rather than the standard t-test. Instead of relying on the t-distribution, students use the class's own data to build a **permutation test** (repeatedly shuffling group labels to see how extreme the observed group difference really is) and a **bootstrap** (resampling within each group to build a confidence interval for the difference in means). Both methods let students see hypothesis testing as a simulation problem rather than a formula lookup, useful groundwork before moving to more complex parametric tests.

## App
Below is an example of what the app looks like when embeded into a Quarto document.
```{=html}
<iframe
  src="Apps/color.html"
  width="100%"
  height="640"
  style="border: none;"
></iframe>
```

## Download
You can download the app directly using the link below. This should open as a standalone app that can open on any device. You can also wrap this around a `=html` code chunk in your Quarto document to achieve a similar effect to what is seen above.
<center>
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(downloadthis)
download_file(
  path = "Apps/color.html",
  button_label = "Colour Game",
  button_type = 'success'
)
```
</center>

## Permutation Test

Once the class's scores are compiled into two groups (Control, Distraction), calculate the observed difference in means:

$$
\hat{\Delta} = \bar{x}_{\text{control}} - \bar{x}_{\text{distraction}}
$$

A permutation test asks: if group membership had no real effect, how often would a difference this large (or larger) show up just from randomly relabelling who was in which group?

```{r, eval=FALSE}
scores <- c(control_scores, distraction_scores)
group  <- c(rep("control", length(control_scores)), rep("distraction", length(distraction_scores)))

observed_diff <- mean(control_scores) - mean(distraction_scores)

n_perm <- 10000
perm_diffs <- numeric(n_perm)

set.seed(1)
for (i in 1:n_perm) {
  shuffled <- sample(group)
  perm_diffs[i] <- mean(scores[shuffled == "control"]) - mean(scores[shuffled == "distraction"])
}

p_value <- mean(abs(perm_diffs) >= abs(observed_diff))
```

Plot a histogram of `perm_diffs` with a vertical line at `observed_diff` — this is a highly visual way for students to see exactly what a p-value represents: the proportion of the shuffled world that looks at least as extreme as what was actually observed.

## Bootstrap Confidence Interval

Rather than testing a hypothesis, the bootstrap estimates a **range of plausible values** for the true difference in means, by resampling *within* each group (with replacement) many times.

```{r, eval=FALSE}
n_boot <- 10000
boot_diffs <- numeric(n_boot)

set.seed(1)
for (i in 1:n_boot) {
  boot_control     <- sample(control_scores, replace = TRUE)
  boot_distraction <- sample(distraction_scores, replace = TRUE)
  boot_diffs[i] <- mean(boot_control) - mean(boot_distraction)
}

ci <- quantile(boot_diffs, c(0.025, 0.975))
```

Compare this bootstrap confidence interval to the observed difference from the permutation test — do they tell a consistent story? Ask the class: does the interval include zero? What would that mean if it did?

## General questions

- Why do we shuffle the *group labels* rather than the scores themselves in a permutation test? What would happen if we shuffled something else?
- The permutation test and the bootstrap both involve "resampling," but they answer different questions. Can you describe, in your own words, what each one is actually trying to tell you?
- If the permutation test's p-value came out just above or just below 0.05, how confident should the class be in that conclusion? What would you want to know before trusting it?
- Why might resampling-based methods like these be more intuitive (or more trustworthy) to a beginner than a formula-based t-test, even though they're often testing a similar hypothesis?
- If you ran the permutation test again with a different random seed, would you expect the p-value to change? By how much, roughly?
- As with the Stack Game, students weren't randomly assigned to Control vs. Distraction. How might that affect how much we can trust either the permutation test or bootstrap result here?