---
title: "Find Your Cluster"
format: html
---

::: {.tag-list}
[K-Means Clustering]{.tag}
[Unsupervised Learning]{.tag}
[Multivariate Data]{.tag}
[Standardisation]{.tag}
:::

## About

This activity uses a short self-report app to give every student a personal profile across six dimensions: **Speed**, **Risk**, **Structure**, **Ambiguity**, **Social**, and **Pattern** — covering things like how quickly they make decisions, how comfortable they are with uncertainty, and whether they lean toward independent or collaborative work. Once the whole class has a six-number profile, the instructor runs k-means clustering live to sort students into groups based on similarity — and if the room allows, students physically move to sit with their cluster, turning an abstract clustering output into something they can immediately feel and discuss.

## Demonstration

## Materials

- The **Six Micro-Tests Score App** (download below) — each student needs a device to access it

<center>
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(downloadthis)
download_file(
  path = "test.html",
  button_label = "Six dimensions test",
  button_type = 'success'
)
```
</center><br>

- A tool for students to submit their six scores (e.g. Google Forms, or directly via the app's "Copy CSV" button if you're collecting responses manually)
- A stats package (R recommended) set up to run k-means clustering live
- Optional: enough open floor space or movable tables for students to physically regroup by cluster afterward

```{r}

```


## The Six Dimensions

| Dimension | What it measures |
|---|---|
| **Speed** | Quick vs. careful decision-making |
| **Risk** | Cautious vs. risk-seeking responses |
| **Structure** | Preference for clear instructions and rules |
| **Ambiguity** | Comfort with unclear situations |
| **Social** | Independent vs. collaborative working style |
| **Pattern** | Rule-based vs. intuitive thinking |

Each test has 5 items on a 1–5 scale (with reverse-scored items handled automatically), giving a raw score from 5–25 per dimension, or a mean item score from 1–5.

## Step-by-Step Plan

### 0 (Pre-tasks)

- Test the app yourself and confirm the "Copy CSV" function works as expected on your setup
- Set up a submission form (or a shared spreadsheet) where students can paste in their name and six scores
- Have your R session or spreadsheet ready with a k-means script prepared in advance (see Step 4), so you're not writing code live in front of the class

### 1 (Setup)

Direct students to the app and explain the task:

> "You're going to complete six very short tests, each with 5 quick questions. Just answer honestly and quickly — there's no right or wrong answer here. Give yourself 5 to 10 minutes to get through all six."

### 2 (Complete the Tests)

Give students 5–10 minutes to work through all six tests in the app. Once finished, they click "Calculate scores" and then "Copy CSV" to get their six-dimension profile.

### 3 (Collect the Data)

Have students paste their name and six scores into your shared submission form or spreadsheet. Compile the full class dataset — each row is one student, each column is one of the six dimension scores.

### 4 (Standardise the Data)

Before running k-means, standardise each dimension so no single test dominates the clustering just because it happens to have more spread.

::: {.callout-note collapse="false"}
## Why standardise first?

K-means clustering is based on distance between points. If one dimension has a much larger range or variance than the others, it will dominate the distance calculation and effectively drown out the other five dimensions. Standardising each column to a mean of 0 and standard deviation of 1 puts all six dimensions on equal footing.

$$
z = \frac{x - \bar{x}}{s}
$$
:::

```r
scores_raw <- read.csv("class_scores.csv")
scores_z <- scale(scores_raw[, c("speed","risk","structure","ambiguity","social","pattern")])
```

### 5 (Choose the Number of Clusters)

Before running k-means, decide how many clusters ($k$) to use. A simple, visual approach for class purposes is the **elbow method**: plot total within-cluster variation against different values of $k$, and look for the point where adding more clusters stops giving a big improvement.

```r
wss <- sapply(1:8, function(k) kmeans(scores_z, centers = k, nstart = 10)$tot.withinss)
plot(1:8, wss, type = "b", xlab = "Number of clusters (k)", ylab = "Within-cluster sum of squares")
```

Discuss with the class where the "elbow" appears to be, and pick a $k$ that's small enough to be meaningful for regrouping the room (e.g. 3–5 clusters).

### 6 (Run K-Means)

```r
set.seed(123)
km <- kmeans(scores_z, centers = 4, nstart = 25)
scores_raw$cluster <- km$cluster
```

Display each student's assigned cluster alongside their name.

### 7 (Interpret the Clusters)

For each cluster, calculate the average (standardised) score on each of the six dimensions, and display it as a table or radar-style summary.

```r
aggregate(scores_z, by = list(cluster = km$cluster), FUN = mean)
```

Ask the class to help name each cluster based on its profile — e.g. a cluster high on Risk and low on Structure might get nicknamed "The Improvisers," while a cluster high on Structure and Pattern might be "The Planners."

### 8 (Move and Discuss)

If the room allows, announce each student's cluster and have them physically move to sit with their group.

Once seated together, give each group a few minutes to discuss:

- Do you feel like this cluster actually describes you? Why or why not?
- What do you have in common with the others at your table, beyond just the six numbers?
- Is there a dimension where your group is surprisingly *split*, even though you were clustered together overall?

## Bonus / Optional:

- Can relate this activity to dimensionality and the curse of high dimensions
- A few extensions to try:
    - **Visualise in 2D:** Run PCA on the standardised data and plot the first two principal components, coloured by cluster, so students can see a 2D "map" of the whole class.
    - **Compare k values:** Re-run the clustering with a different $k$ (e.g. 3 vs. 6) and discuss how the groupings — and the physical seating — would have changed.
    - **Hierarchical clustering:** Run an agglomerative hierarchical clustering on the same data and display a dendrogram, comparing which students get grouped together versus the k-means result.
    - **Stability check:** Re-run k-means a few times with different random starting points (`nstart`) and discuss why the algorithm can converge to different results depending on initialisation — a nice segue into why `nstart` with multiple restarts is standard practice.